A Bayesian networks approach to infer social changes from burials in northeastern Taiwan during the European colonization period

### 2021-08-21 analysis/paper/paper.Rmd
knit and resubmit to JAS
edit the proof after acceptance

---
title: "A Bayesian networks approach to infer social changes in burials in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to infer interactions and relationships between entities. Statistical approaches to network analysis, such as exponential random graph models (ERGMs), provide a way to test hypotheses about dynamic processes of network formation. However, computational difficulties and sensitivity to uncertainties limit the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification of uncertainty, and robust model evaluation of network properties. We tested a hypothesis of social change relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition among the burials from network ties based on ritual objects to wealth objects, and a more centralized structure with increased social differentiation after the European presence was established in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  1. We present Bayesian inference on exponential random graph models (ERGMs) to model archaeological networks.
  2. We explore social processes and network structures in archaeological data from two time periods.
  3. Differences in network statistics imply change over time in social structures influenced by foreign presence. 
  4. Bayesian ERGMs are efficient and useful for evaluating and interpreting networks. 
---

@@

<!-- ERGMs in archaeology--> 
Exponential random graph models (ERGMs) are statistical models for social networks that allows direct modeling for the formation of edges, or ties, between nodes [@Robins2007]. The application of ERGMs first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites and at 190 sites for a later study [@Brughmans2015]. By fitting models with the observed network of archaeological data, they proposed that ERGMs are a promising approach for exploration and hypothesis testing of social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. Other examples, such as exploring drivers of social stability for the Valencian Bronze Age settlements in eastern Spain [@Cegielski2020], or reconstructing structurally efficient networks of Middle Bronze Age settlements in the Aegean [@Amati2018], demonstrate the usefulness of ERGMs for hypothesis testing for social processes.

@@

<!--Intro to our case study and model-->
We use a Bayesian ERGM to study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan (Figure \@ref(fig:figure-KWL-map)), which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally an influx of Han Chinese in the 19th century [@Chen2007]. Our anthropological model proposes that the influence of an indirect colonial power, combined with high local values attached to imported goods, led to increased social inequality due to competition among individuals (details in Supplementary Online Materials) [@Brumfiel1994; @Clark1994]. Before the arrival of Europeans, some imported goods already entered Kiwulan because of the regional trade network with Chinese traders [@Chen2005; @Cheng2017]. We assume that the use of imported goods and their values were amplified after the European arrival. Different from Heping dao and Tamsui in northern Taiwan where European forts are located, the remote geographical setting of northeastern Taiwan made it less interfered by European colonial power but experienced frequent trade activities due to local natural resources, such as rice and deerskins [@Chen2005; @Cheng2017]. This makes northeastern Taiwan a special area to explore the effect of European colonization in a pericolonial context that is still not well understood compared to European colonial bases [@Acabado2017; @Trabert2017].

<!--theoretical framework: corporate/network strategies-->
The involvement in long-distance exchange is often associated with an increased social inequality where ambitious individuals could gain power by building personal influence through the acquisition and distribution of foreign goods, leading to social change from corporate to network societies [@Blanton1996; @Feinman2000; @Klehm2017]. Corporate-network societies represent two distinct political-economic strategies for obtaining power, viewed as a horizontal axis crosscutting traditional vertical hierarchy [@Drennan2010; @Feinman2000]. The features of network strategies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks. These are contrasted with corporate strategies that stress communal ritual elements, shared power, and where wealth differentiation, if any, would be associated with corporate groups, such as age-based or kin-based, rather than individuals [@Siegel1999]. In Asia, network-based societies are frequently identified, especially for places involved in long-distance trade [@Carter2015; @Ueda2016; @Liu2006].

@@

<!--hypothesis and research questions-->
Using the theoretical framework of corporate-network strategies, we test a hypothesis of the shift in organizational modes in relation to increased social inequality, from a more corporate to a more network society, after the European presence that can be observed in the Kiwulan burial network [@Blanton1996; @Drennan2010; @Feinman2000]. We explore whether this change can be identified through the pattern of intercorrelations of foreign prestige goods among burials. We ask: (1) Do the burial network after European arrival present a more centralized structure with fewer subgroups compared to the burial network before European contact? and if so, (2) what are the major variables affecting or forming these two networks? Can we identify the shift in variables from ritual to wealth associated with two different strategies? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

<!-- Kiwulan burials and a review of current studies of social organization -->
We analyzed burial data collected from the excavation reports, and the original fieldwork notes for the upper component of Kiwulan (1350-1850 AD) [@Chen2007]. A total of 90 burials were unearthed from adjacent excavation squares that provide continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). The preservation of human remains at Kiwulan was generally poor, so we have incomplete data of age and sex (details in Supplementary Online Materials) and lack health status data. Burials are oriented in an east-west direction on the north side of the residential area, which is indicated by post-holes and _in-situ_ wooden posts, suggesting a well-organized spatial arrangement of houses. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on a comparative analysis of the frequencies of all burial goods. She found that the burials with rare prestige goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing behaviors from different phases. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period (before European colonization) to the post-European period (during and after European colonization until the arrival of Han Chinese immigrants).

@@

<!--assigning burials based on our chronology-->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [for more details see @LWandBM2020ornament; @LWandBM2020pot]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. For the two periods studied here, we evaluate differences in network size were using a vertex bootstrap method. We also excluded `r disturb` burials that were heavily disturbed by modern construction which prevented accurate determination of their chronology. Most burials have clear stratigraphic contexts so modern or ancient disturbance is readily visible (i.e. if a burial has been cut into by later burials, very few cases of this have been documented at Kiwulan).  By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

```{r raincloud-beads, fig.cap = "The distributions of different types of trade beads across all burials analyzed in this paper."}
knitr::include_graphics(here::here("analysis", "figures", "000-raincloud-beads.png"))
```

@@

```{r table-ergm-terms}
ergm_terms <- readxl::read_excel(here::here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
ergm_terms_selected <-
  ergm_terms %>% 
  select(-"Description", -"note")
knitr::kable(ergm_terms_selected,
             caption = "Network parameters in ERGMs used for model specifications with associated archaeological interpretation for burial relations. ERGM terminology is from the R package statnet (Handcock, 2008; Morris, 2008)")
```

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Priors are a distinctive aspect of Bayesian methods that enable us combining our beliefs based on prior research with the newly observed data [@Gill2014]. The normal distribution N($\mu$, $\sigma^2$), where $\mu$ is the mean and $\sigma^2$ is the standard deviation, is typically used for priors when modelling network parameters because it is sensible and convenient, allowing adjustments for a wide range of prior information [@Caimo2017; @Normand1992]. The larger the mean of the prior distribution, the more positive effect of that parameter on network formation. A larger standard deviation makes the prior more vague, enabling comparison between variables for which we do not have specific assumptions. Since low network densities are commonly found in the real world [@Caimo2017], we specified the prior of the edge density parameter to low for both network models, following a normal distribution with mean at `r priors$mean_same[1]`, and standard deviations at `r priors$sigma_same[1]` (i.e (N(`r priors$mean_same[1]`, `r priors$sigma_same[1]`))).

To evaluate our anthropological model of increased social inequality, specifically a shift from a corporate to network society [@Drennan2010; @Feinman2000], we specified priors that are meaningful for social inequality, especially for the variables of transitivity and centralization. For the network before European contact, we set the priors to model less social differentiation and emphasize the ritual element shared in corporate groups: higher transitivity (N(`r priors$pre_prior_mean[6]`, `r priors$pre_prior_sigma[6]`)), lower centralization (N(`r priors$pre_prior_mean[7]`, `r priors$pre_prior_sigma[7]`)), and a higher covariate effect of ritual activity (N(`r priors$pre_prior_mean[4]`, `r priors$pre_prior_sigma[4]`)). Conversely, to model an increased social differentiation after European contact, we set the priors for the network after European arrival to lower transitivity (N(`r priors$post_prior_mean[6]`, `r priors$post_prior_sigma[6]`)), higher centralization (N(`r priors$post_prior_mean[7]`, `r priors$post_prior_sigma[7]`)), and a higher covariate effect for burial values (N(`r priors$post_prior_mean[5]`, `r priors$post_prior_sigma[5]`)).  For covariates about biological features, such as age, sex, we specified a vague prior (N(`r priors$mean_same[2]`, `r priors$sigma_same[2]`)) for both models to explore their effects. Similarly, we used the same vague prior (N(`r priors$mean_same[8]`, `r priors$sigma_same[8]`)) for physical distance between burials, to explore whether there is kinship-based proximity, e.g. stronger correlations for shorter distances. We then examine the fit of the observed data to the two models we specified to evaluate our hypotheses.

@@

```{r table-posterior-stats}
# tidy data for the table of posterior stats 
posterior_esti_two_phases<-
  bergm_two_phases %>%
  pivot_longer(!phase, 
               names_to = "parameter", 
               values_to = "posterior") %>%
  mutate(parameter= case_when(parameter == "edges" ~ "density",
                              parameter == "nodematch.age" ~ "age-homophily",
                              parameter == "nodematch.sex" ~ "sex-homophily",
                              parameter == "nodematch.ritual" ~ "ritual-homophily",
                              parameter == "nodematch.value" ~ "wealth-homophily",
                              parameter == "gwesp" ~ "transitivity",
                              parameter == "gwdeg" ~ "centralization",
                              parameter == "dyadcov.distance" ~ "physical distance")) %>% 
  mutate(parameter = factor(parameter,
                            levels=c("density", "age-homophily", 
                                     "sex-homophily", "ritual-homophily",
                                     "wealth-homophily", "transitivity",	
                                     "centralization", "physical distance"))) %>% 
  group_by(parameter, phase) %>% 
  summarize(mean = round(mean(posterior), 2),
            median = round(quantile(posterior, 0.5), 2),
            `2.5%` = round(quantile(posterior, 0.025), 2),
            `97.5%` = round(quantile(posterior, 0.975), 2)) %>% 
  pivot_wider(names_from = phase, 
              values_from = c(mean, median, `2.5%`, `97.5%`)) %>% 
  select(parameter, ends_with(c("pre", "post")))

@@

```{r figure-posterior-distribution, fig.cap = "Posterior density estimates for the parameters associated with edges, transitivity, centralization, and physical distance by phases. The pre-European group presents a long tailed distribution with larger values for the transitivity, but negative values for the centralization parameter compared to the post-European group. The edges and distance parameters for the pre-European group have larger values generally compared to pos-European one."}
knitr::include_graphics(here::here("analysis", "figures", "006-posterior-distribution.png"))
```

@@

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To confirm the robustness of our comparison between these two different-sized networks, we used a vertex bootstrap technique to cross-validate the results of our Bayesian ERGMs. The vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. nodes) to quantify standard errors and estimate sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks of different sizes by examining their confidence intervals for the network population. We used the vertex bootstrap to compute endogenous network statistics, including density, centralization, and transitivity for our two networks. We explored the sample size effect by removing nodes at certain percentages (5-40%) for both networks and comparing their confidence intervals. The results demonstrate a consistent difference between the two networks up to 35% node removals for the centralization variable as shown in Figure \@ref(fig:figure-bootstrap-CI). This shows consistency with our earlier finding of negative centralization in the pre-European period and positive centralization in the post-European period, over several conditions of node removal. Similarly consistent with our earlier result, the vertex bootstrap shows no significant difference on density and transitivity. This indicates that the network variables, including density, centralization, and transitivity, are robust to the different network sizes of our sample.

@@

<!--archaeological implication based on the results-->
Network covariates give insights into the socio-cultural factors that were associated with the acquisition and distribution of foreign prestige goods. The general pattern of posterior distributions indicates that the main factor contributing to burial networks for the pre-European network was ritual, and for the post-European network was wealth. We interpret these results as illustrating different social and economic mechanisms influencing relations between burials for the two periods. Before European arrival, ritual behaviors could have been important factors that structured the interconnections of prestige goods in the burial contexts. This might imply that ritual was status indicator at the time before many foreign goods were introduced to Indigenous societies. After the European presence was established, status indicators shifted from ritual to wealth differences, corresponding with the characteristics of corporate/network strategies. In addition, geographical proximity of burials, implying kinship association, has some effect on sharing similar prestige items between burials in the pre-European period, but not in the post-European period. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based after European arrival. Burials in the same age group tend to have similar goods after European arrival. However the sex of the buried individual did not have any effects on network formation for both networks. We should also note that their effects may be impacted by the unavailability of age (`r age_miss` out of `r pre_and_post` burials) and sex (`r sex_miss` out of `r pre_and_post` burials) values in our sample.

@@

<!-- trade dynamic during this period and future work for comparison -->
We recognize that complex local and international trade dynamics between different Indigenous groups and other foreign presence, Han Chinese and Japanese, might have contributed to social changes at Kiwulan as well. However, the current changes observed from burial networks and spatial distribution of trade items in the residential area at Kiwulan demonstrate consistent changes before and after the European arrival. The frequent trade activities stimulated by the presence of Europeans led to many trade items, including Anping jars, stonewares, Chinese porcelains, and beads found at archaeological sites with Dutch occupation at different parts of Taiwan, such as Fort Zeelandia. As a large village with continuous occupation since the 14th century in northeastern Taiwan, Kiwulan provides a special example to explore the reaction of Indigenous people living on the periphery to European colonial presence at a settlement scale. Was its experience with the European presence unique or typical of the region? We need more evidence from other sites in northeastern Taiwan to dive deeper into pericolonial effects, and compare evidence between sites on the periphery and colonial centers, such as Heping Dao and Tamsui. Future work should focus on collecting data from Indigenous sites related to European contact for comparison with Kiwulan.

>>>>

---
title: "A Bayesian networks approach to infer social changes from burials in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to infer interactions and relationships between entities. Statistical approaches to network analysis, such as exponential random graph models (ERGMs), provide a way to test hypotheses about dynamic processes of network formation. However, computational difficulties and sensitivity to uncertainties limit the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification of uncertainty, and robust model evaluation of network properties. We tested a hypothesis of social change relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition among the burials from network ties based on ritual objects to wealth objects, and a more centralized structure with increased social differentiation after the European presence was established in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social inequality; Taiwan
highlights: |
  1. We present Bayesian inference on exponential random graph models (ERGMs) to model archaeological networks.
  2. We explore social processes and network structures in archaeological data from two time periods.
  3. Differences in network statistics imply change over time in social structures influenced by foreign presence. 
  4. Bayesian ERGMs are efficient and useful for evaluating and interpreting networks. 
---

@@

<!-- ERGMs in archaeology--> 
Exponential random graph models (ERGMs) are statistical models for social networks that allow direct modeling for the formation of edges, or ties, between nodes [@Robins2007]. The application of ERGMs first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites and at 190 sites for a later study [@Brughmans2015]. By fitting models with the observed network of archaeological data, they proposed that ERGMs are a promising approach for exploration and hypothesis testing of social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. Other examples, such as exploring drivers of social stability for the Valencian Bronze Age settlements in eastern Spain [@Cegielski2020], or reconstructing structurally efficient networks of Middle Bronze Age settlements in the Aegean [@Amati2018], demonstrate the usefulness of ERGMs for hypothesis testing for social processes.

@@

<!--Intro to our case study and model-->
We use a Bayesian ERGM to study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan (Figure \@ref(fig:figure-KWL-map)), which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally an influx of Han Chinese in the 19th century [@Chen2007]. Our anthropological model proposes that the influence of an indirect colonial power, combined with high local values attached to imported goods, led to increased social inequality due to competition among individuals (details in Supplementary Online Materials) [@Brumfiel1994; @Clark1994]. Before the arrival of Europeans, some imported goods already entered Kiwulan because of the regional trade network with Chinese traders [@Chen2005; @Cheng2017]. We assume that the use of imported goods and their values were amplified after the European arrival. Different from Heping dao and Tamsui in northern Taiwan where European forts are located, the remote geographical setting of northeastern Taiwan resulted in less interference from European colonial powers, but more frequent trade activities due to local natural resources, such as rice and deerskins [@Chen2005; @Cheng2017]. This makes northeastern Taiwan a special area to explore the effect of European colonization in a pericolonial context that remains not well understood compared to European colonial bases [@Acabado2017; @Trabert2017].

<!--theoretical framework: corporate/network strategies-->
Involvement in long-distance exchange is often associated with an increased social inequality where ambitious individuals could gain power by building personal influence through the acquisition and distribution of foreign goods, leading to social change from corporate to network societies [@Blanton1996; @Feinman2000; @Klehm2017]. Corporate-network societies represent two distinct political-economic strategies for obtaining power, viewed as a horizontal axis crosscutting traditional vertical hierarchy [@Drennan2010; @Feinman2000]. The features of network strategies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks. These are contrasted with corporate strategies that stress communal ritual elements, shared power, and where wealth differentiation, if any, would be associated with corporate groups, such as age-based or kin-based, rather than individuals [@Siegel1999]. In Asia, network-based societies are frequently identified, especially for places involved in long-distance trade [@Carter2015; @Ueda2016; @Liu2006].

@@

<!--hypothesis and research questions-->
Using the theoretical framework of corporate-network strategies, we test a hypothesis of the shift in organizational modes in relation to increased social inequality, from a more corporate to a more network society, after the European presence that can be observed in the Kiwulan burial network [@Blanton1996; @Drennan2010; @Feinman2000]. We explore whether this change can be identified through the pattern of intercorrelations of foreign prestige goods among burials. We ask: (1) Does the burial network after European arrival present a more centralized structure with fewer subgroups compared to the burial network before European contact? and if so, (2) what are the major variables affecting or forming these two networks? Can we identify the shift in variables from ritual to wealth associated with two different strategies? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

<!-- Kiwulan burials and a review of current studies of social organization -->
We analyzed burial data collected from the excavation reports, and the original fieldwork notes for the upper component of Kiwulan (1350-1850 AD) [@Chen2007]. A total of 90 burials were unearthed from adjacent excavation squares that provide continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). The preservation of human remains at Kiwulan was generally poor, so we have incomplete data for age and sex (details in Supplementary Online Materials) and lack health status data. Burials are oriented in an east-west direction on the north side of the residential area, which is indicated by post-holes and _in-situ_ wooden posts, suggesting a well-organized spatial arrangement of houses. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on a comparative analysis of the frequencies of all burial goods. She found that the burials with rare prestige goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing behaviors from different phases. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period (before European colonization) to the post-European period (during and after European colonization until the arrival of Han Chinese immigrants).

@@

<!--assigning burials based on our chronology-->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [for more details see @LWandBM2020ornament; @LWandBM2020pot]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size that corresponds to the population decline at Kiwulan in the Chinese period [@Chen2007]. For the two periods studied here, we evaluated differences in network size using a vertex bootstrap method. We also excluded `r disturb` burials that were heavily disturbed by modern construction which prevented accurate determination of their chronology. Most burials have clear stratigraphic contexts so modern or ancient disturbance is readily visible (i.e. if a burial has been cut into by later burials, very few cases of this have been documented at Kiwulan).  By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

```{r raincloud-beads, fig.cap = "The distributions of different types of trade beads across all burials analyzed in this paper. Seven burials with 60 and more (the maximum is 2936) glass beads are not shown in this plot."}
knitr::include_graphics(here::here("analysis", "figures", "000-raincloud-beads.png"))
```

@@

```{r table-ergm-terms}
ergm_terms <- readxl::read_excel(here::here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
ergm_terms_selected <-
  ergm_terms %>% 
  select(-"Description", -"note")
knitr::kable(ergm_terms_selected,
             caption = "Network parameters in ERGMs used for model specifications with associated archaeological interpretation for burial relations. ERGM terminology is from the R package statnet (Handcock et al., 2008; Morris et al., 2008)")
```

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Priors are a distinctive aspect of Bayesian methods that enable us to combine our beliefs based on prior research with newly observed data [@Gill2014]. The normal distribution N($\mu$, $\sigma^2$), where $\mu$ is the mean and $\sigma^2$ is the standard deviation, is typically used for priors when modelling network parameters because it is sensible and convenient, allowing adjustments for a wide range of prior information [@Caimo2017; @Normand1992]. The larger the mean of the prior distribution, the more positive effect of that parameter on network formation. A larger standard deviation makes the prior more vague, enabling comparison between variables for which we do not have specific assumptions. Since low network densities are commonly found in the real world [@Caimo2017], we specified the prior of the edge density parameter to low for both network models, following a normal distribution with mean at `r priors$mean_same[1]`, and standard deviations at `r priors$sigma_same[1]` (i.e (N(`r priors$mean_same[1]`, `r priors$sigma_same[1]`))).

To evaluate our anthropological model of increased social inequality, specifically a shift from a corporate to network society [@Drennan2010; @Feinman2000], we specified priors that are meaningful for social inequality, especially for the variables of transitivity and centralization (see our supplementary information for more details on prior selection). For the network before European contact, we set the priors to model less social differentiation and emphasize the ritual element shared in corporate groups: higher transitivity (N(`r priors$pre_prior_mean[6]`, `r priors$pre_prior_sigma[6]`)), lower centralization (N(`r priors$pre_prior_mean[7]`, `r priors$pre_prior_sigma[7]`)), and a higher covariate effect of ritual activity (N(`r priors$pre_prior_mean[4]`, `r priors$pre_prior_sigma[4]`)). Conversely, to model an increased social differentiation after European contact, we set the priors for the network after European arrival to lower transitivity (N(`r priors$post_prior_mean[6]`, `r priors$post_prior_sigma[6]`)), higher centralization (N(`r priors$post_prior_mean[7]`, `r priors$post_prior_sigma[7]`)), and a higher covariate effect for burial values (N(`r priors$post_prior_mean[5]`, `r priors$post_prior_sigma[5]`)).  For covariates about biological features, such as age, sex, we specified a vague prior (N(`r priors$mean_same[2]`, `r priors$sigma_same[2]`)) for both models to explore their effects. Similarly, we used the same vague prior (N(`r priors$mean_same[8]`, `r priors$sigma_same[8]`)) for physical distance between burials, to explore whether there is kinship-based proximity, e.g. stronger correlations for shorter distances. We then examine the fit of the observed data to the two models we specified to evaluate our hypotheses.

@@

```{r table-posterior-stats}
# tidy data for the table of posterior stats 
posterior_esti_two_phases<-
  bergm_two_phases %>%
  pivot_longer(!phase, 
               names_to = "parameter", 
               values_to = "posterior") %>%
  mutate(parameter= case_when(parameter == "edges" ~ "Density",
                              parameter == "nodematch.age" ~ "Age-homophily",
                              parameter == "nodematch.sex" ~ "Sex-homophily",
                              parameter == "nodematch.ritual" ~ "Ritual-homophily",
                              parameter == "nodematch.value" ~ "Wealth-homophily",
                              parameter == "gwesp" ~ "Transitivity",
                              parameter == "gwdeg" ~ "Centralization",
                              parameter == "dyadcov.distance" ~ "Physical distance")) %>% 
  mutate(parameter = factor(parameter,
                            levels=c("Density", "Age-homophily", 
                                     "Sex-homophily", "Ritual-homophily",
                                     "Wealth-homophily", "Transitivity",	
                                     "Centralization", "Physical distance"))) %>% 
  group_by(parameter, phase) %>% 
  summarize(mean = round(mean(posterior), 2),
            median = round(quantile(posterior, 0.5), 2),
            `2.5%` = round(quantile(posterior, 0.025), 2),
            `97.5%` = round(quantile(posterior, 0.975), 2)) %>% 
  pivot_wider(names_from = phase, 
              values_from = c(mean, median, `2.5%`, `97.5%`)) %>% 
  select(parameter, ends_with(c("pre", "post")))

@@

```{r figure-posterior-distribution, fig.cap = "Posterior density estimates for the parameters associated with edges, transitivity, centralization, and physical distance by phases. The pre-European group presents a long tailed distribution with larger values for the transitivity, but negative values for the centralization parameter compared to the post-European group. The edges and distance parameters for the pre-European group have larger values generally compared to the post-European one."}
knitr::include_graphics(here::here("analysis", "figures", "006-posterior-distribution.png"))
```

@@

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To confirm the robustness of our comparison between these two different-sized networks, we used a vertex bootstrap technique to cross-validate the results of our Bayesian ERGMs. The vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. nodes) to quantify standard errors and estimate sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks of different sizes by examining their confidence intervals for the network population. We used the vertex bootstrap to compute endogenous network statistics, including density, centralization, and transitivity for our two networks. We explored the sample size effect by removing nodes at certain percentages (5-40%) for both networks and comparing their confidence intervals. The results demonstrate a consistent difference between the two networks at up to 35% node removals for the centralization variable as shown in Figure \@ref(fig:figure-bootstrap-CI). This shows consistency with our earlier finding of negative centralization in the pre-European period and positive centralization in the post-European period, over several conditions of node removal. Similarly consistent with our earlier result, the vertex bootstrap shows no significant difference on density and transitivity. This indicates that the network variables, including density, centralization, and transitivity, are robust to the different network sizes of our sample.

@@

<!--archaeological implication based on the results-->
Network covariates give insights into the socio-cultural factors that were associated with the acquisition and distribution of foreign prestige goods. The general pattern of posterior distributions indicates that the main factor contributing to burial networks for the pre-European network was ritual, and for the post-European network was wealth. We interpret these results as illustrating different social and economic mechanisms influencing relations between burials for the two periods. Before European arrival, ritual behaviors could have been important factors that structured the interconnections of prestige goods in the burial contexts. This might imply that ritual was a status indicator at the time before many foreign goods were introduced to Indigenous societies. After the European presence was established, status indicators shifted from ritual to wealth differences, corresponding with the characteristics of corporate/network strategies. In addition, geographical proximity of burials, implying kinship association, has some effect on sharing similar prestige items between burials in the pre-European period, but not in the post-European period. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based after European arrival. Burials in the same age group tend to have similar goods after European arrival. However the sex of the buried individual did not have any effects on network formation for both networks. We should also note that their effects may be impacted by the unavailability of age (`r age_miss` out of `r pre_and_post` burials) and sex (`r sex_miss` out of `r pre_and_post` burials) values in our sample.

@@

<!-- trade dynamic during this period and future work for comparison -->
We recognize that complex local and international trade dynamics between different Indigenous groups and other foreign groups, such as the Han Chinese and Japanese, might have contributed to social changes at Kiwulan as well. However, the current changes observed from burial networks and spatial distribution of trade items in the residential area at Kiwulan demonstrate consistent changes before and after the European arrival. The frequent trade activities stimulated by the presence of Europeans led to many trade items, including Anping jars, stonewares, Chinese porcelains, and beads found at archaeological sites with Dutch occupation at different parts of Taiwan, such as Fort Zeelandia. Although shortages mentioned in Spanish documents might indicate insufficient goods were brought into Taiwan [@Borao2009], we observed a large amount of trade goods at Kiwulan. Trade goods may not have been originally owned by Europeans if we consider the pre-existent regional trade with Chinese merchants. However, the large amount of foreign goods suggests frequent and active trade activities with local places involved in a global trade network.

<!-- future work -->
As a large village with continuous occupation since the 14th century in northeastern Taiwan,  the archaeological evidence at Kiwulan provides a special example to explore the reaction of Indigenous people living on the periphery to European colonial presence at a settlement scale that historical records rarely offer. Was its experience with the European presence unique or typical of the region? We need more evidence from other sites in northeastern Taiwan to dive deeper into pericolonial effects, and compare evidence between sites on the periphery and colonial centers, such as Heping Dao and Tamsui. The contemporary community of descendants of Kiwulan, the Kavalan people, shows features of a corporate society, such as a matrilineal system with age group administration and ritual practices [@Liu2021]. This indicates that social change has been continuous in this area, from the more centralized network mode of social organization we observed in the burials after European arrival, to the current corporate model of social organization. Future work should focus on collecting data from Indigenous sites related to European contact and Chinese contact for comparison with Kiwulan.
